PEACE: {\underline P}arallel {\underline E}nvironment for {\underline A}ssembly and {\underline C}lustering of Gene {\underline E}xpression

### 2010-02-12 papers/NAR_web_server_2009/paper.tex
Incorperation of comments on manuscript from Mufit and Chun

\begin{abstract}
  We present {\sc Peace}, a tool for the high-throughput clustering of
  transcript fragment sequences produced by Next Generation or Sanger
  Sequencing technologies.  Installed and run through an easily
  downloaded GUI, {\sc Peace} can process large data sets, grouping the
  fragments by gene association, faster and with greater sensitivity
  than competing clustering tools such as {\sc WCD}, and with
  considerably greater sensitivity than the implicit clustering
  performed by assembly tools such as {\sc Cap3}.  Through the GUI the
  user can collect cluster statistics and examine specific
  clusters for more comprehensive study.  And by using {\sc Peace} as
  a ``pre-assembly'' tool to allow the application of an assembly tool
  to the resultant individual clusters, the user gains both a significant
  increase in sensitivity of the clustering method and a large decrease
  in overall runtime for the clustering and assembly together.
\end{abstract}

@@

Understanding an organism's transcriptome, the set of (spliced)
transcripts expressed by genes of the organism, is a vital step in
understanding the full functional and organizational role of the genome
in the life cycle of any organism.  Studying the transcriptome has led
to gene discovery, provided information on protein isoforms, and
generally helped shed light on the biological processes both
controlling and dictated by the genome \cite{Nagaraj07}.  But in order
to get access to those transcripts, we must first deal with the
fragmented data produced by both Next Generation and traditional
Sanger sequencing technology.

Until recently, a primary method of gaining access to a transcriptome
sequence was through the use of Expressed Sequence Tags (ESTs), which
were primarily single-pass cDNA sequences derived from transcribed
mRNAs and traditionally sequenced by Sanger Sequencing technology.
More recently, Next-Generation Sequencing (NGS) technologies have
replaced the Sanger Sequencing techniques, allowing for considerably
greater depth and coverage of the transcriptome.  For example, ESTs in
GenBank dbEST are increasingly the product of NGS technologies such as
454 pyrosequencing, which enables the sequencing of novel and rate
transcripts at a considerably higher rate of coverage than did Sanger
Sequencing \cite{Cheung2006,Emrich2007}.  But from a computational
perspective, this is a mixed blessing: while NGS provides immense
quantities of new information, it also provides immensely larger data sets,
requiring even more efficient algorithms than were already needed
for analysis.

Given a set of transcript fragments, produced by either sequencing
technology, sampled from multiple transcripts across the genome, a
necessary first step is that of clustering: separating the fragments
according to the transcript from which they were derived.  While
clustering is frequently performed implicitly by assembly tools such
as {\sc Cap3} \cite{Huang99}, clustering the data as a
``pre-assembly'' step has a number of advantages.  Most significant
among these: performing this step will allow the application of the
assembly tool to individual clusters -- saving significant amounts of
time due to the reduced input size \cite{Hazelhurst08a}.  But
clustering is a computationally challenging problem, especially when
done {\it ab initio} instead of relying on access to a sequenced
genome.  Even with the smaller number of ESTs produced using Sanger
sequencing, the runtime and memory requirements to cluster based on
pair-wise sequence alignments make such an approach infeasible.  The
much larger data-set size produced by next-generation sequencing
technologies exacerbates this problem.

Here we present {\sc Peace}, a tool for the {\it ab initio} clustering
of transcript fragments by gene, applicable to both short-read and traditional
Sanger Sequencing technologies.  Available through the
\href{http://www.peace-tools.org}{www.peace-tools.org} website, the PEACE GUI allows the user
to both easily install (locally or remotely) and run the computational
tool, as well as providing various tools for result analysis. {\sc
  Peace} can be run in single or multi-processor mode, and requires no
Internet connectivity past the initial download.  Using a {\it minimum
  spanning tree} (MST) based clustering strategy \cite{Jain99,Wan08}
and the $d^2$ sequence distance function \cite{Hide94}, {\sc Peace}
generates the MSTs using a tailored version of Prim's algorithm
\cite{Prim57}.  The result is an easy-to-use clustering tool that is considerably
more sensitive to sequencing errors than the competing {\sc wcd} tool
\cite{Hazelhurst08a} and others in the literature without any
sacrifice of performance (e.g. \cite{Burke99}, \cite{Slater00},
\cite{Huang99}, \cite{Parkinson02}, \cite{Kalyanaraman03},
\cite{Malde03}, \cite{Ptitsyn05}, \cite{Picardi09}).

@@

\section{PEACE: Installation and Use}

The {\sc Peace} GUI, available from the {\sc Peace} website
(\href{http://www.peace-tools.org}{www.peace-tools.org}), handles the
details of both installation and job management and provides tools for
the initial analysis of the results.  By downloading and executing the
{\tt peace.jar} GUI Java program on a Linux, Windows or OS-X machine
running Java 1.6+, the user is presented with the {\sc Peace}
management tool.  Through this the user can install the computational
tool and perform a clustering of a data file, view an initial analysis
of the clusters and produce files containing subsets of the clusters
for assembly by a tool such as {\sc Cap3} \cite{Huang99}.  A typical
(first) use of the tool would generally require the the following
steps.

\noindent {\bf Tool Installation:} Installing the GUI requires only
that the Java file be downloaded from the website.  To install the {\sc
  Peace} tool onto a local or remote machine, the user will select
from within the GUI the {\tt Server > Add New Server} tab from the GUI
main menu (Figure~1(a)), which then starts the install
wizard that will prompt for the appropriate information.
Figure~1(b) illustrates the request for server information;
in this example, the user has chosen to install the {\sc Peace}
computational tool on a remote machine and is providing the necessary
connection information.  Server information is persistent between GUI
sessions, giving the user access to {\sc Peace} on the target machine
as needed.

\noindent {\bf Job Processing:} After importing the target file into
the GUI ({\tt File > New Data Set}), the user can start a new job with
the {\tt Job > Job to Complete Clusters} tab and follow the wizard
menus.  Figure~1(c) illustrates the process of specifying
the number of processors available (if running on a machine supporting
the OpenMPI protocol -- which will be determined during job
instillation).  Once executed, the GUI will manage the job thread,
alert the user when the job is completed (or when the user next
runs the GUI after completion) and 
copy the final results back to the local machine if necessary.

\noindent {\bf Result Analysis:} Once the resulting clusters have been
computed and (if necessary) copied back, the user has several options
for analysis:
\begin{itemize}
\item {\bf Export:} The user can export the contents of one or more
  clusters into a FASTA format file, obtaining a subset of the original file
  containing the sequences corresponding to the selected clusters
  ready for processing by an assembly tool (e.g. {\sc Cap3} \cite{Huang99}).
\item {\bf View Clustering}: The user may view a list of all clusters,
  expanding selected clusters to a list of all fragments (illustrated in
  Figure~1(d)).  
\item {\bf Classified Summary Graph:} The user may view a distribution
  of cluster sizes in a linear or $\log$ scale.  Further, the user may
  set up a {\it classifier}, associating certain patterns with
  specific colors.  These patterns will be matched against the
  fragment header information from the original FASTA file, allowing
  the overlay of a colored cluster size distribution does clusters
  containing matching elements in a specified colors (comment by molerjc: please reword the preceding sentence fragment).  For example, if
  the sequence names contain unique string patterns denoting different
  cDNA libraries, the classifier can help the user to determine the
  differential expression profiles of different libraries for a given
  cluster by color coding the strains.  The method of setting up a
  these classifiers, and the resulting histogram, is illustrated in
  Figure~1(e).
\end{itemize}

Extensive documentation for the tool has been posted on the {\sc
  Peace} website, as well as links to several videos demonstrating
{\sc Peace} use and capabilities.

@@

We have tested {\sc Peace} on both simulated and real data,
comparing results against those produced by the {\sc wcd}
clustering tool \cite{Hazelhurst08a} and the {\sc Cap3} assembly
tool \cite{Huang99} (which calculates a cluster as an intermediate
step).  For our simulation tests we used the {\bf ESTSim} tool
\cite{Hazelhurst03} to generate simulated transcript fragments under varying models
of error (Supplementary Materials, Section~\href{ESTsim_sup}),
generating the fragments from the list of 100 zebra fish genes used in
the {\sc wcd} testing \cite{Hazelhurst08a}.  Tool parameters were
taken to match, as closely as possible, those used in the {\sc wcd}
study (see supplementary materials).  Our primary methods of quality
assessment were {\it sensitivity} (the fraction of fragment pairs from the
same gene that were correctly clustered together) and {\it Type 1
  error} (the fraction of genes that were divided between clusters)
\cite{Wang04,Hazelhurst08a}; other measurements are also discussed
in the supplementary materials.  In Figure~2(left) we see a
significant improvement of {\sc Peace} (blue) over {\sc wcd}
(green) and {\sc Cap3} (black) in its sensitivity to sequencing
errors, while in Figure~2(right) we see a comparable
improvement in Type 1 error.  In Figure~3 we look at
the number of singleton clusters (fragments not joined to any cluster),
which will not occur in our simulated sets; we again see significant
improvement in {\sc Peace}.  The sequential runtime of
{\sc Peace} is slightly improved over that of {\sc wcd} (see
Supplementary Materials, Figure~S4).

\begin{figure*}[b]
\centerline{
\includegraphics[trim=0cm 3cm 0cm 1cm, clip, scale=0.9]{screen.d/screen.pdf} 
}
\label{screen}
\caption{Screenshots of the PEACE GUI during execution, including (a)
GUI Welcome and server installation menu; (b) setup wizard for
installing the computational tool on a remote server; (c) execution
wizard for starting a selected job to be executed in parallel mode;
(d) basic cluster output; and (e) histogram view of cluster results
and classifier editor for setting up differential expression profiles.}
\end{figure*}

@@

In applying the tools to real data, we started with a set of
approximately $190,000$ clean Sanger ESTs derived from the {\it
  Chlamydomonas reinhardtii} genome \cite{Liang2008}.  To compute
sensitivity we used the {\it gmap} tool \cite{Wu05} to map the
fragment set to the genome, taking this as our reference clustering.
We see slight improvements in {\sc Peace} over {\sc wcd} in both
sensitivity and Type 1 error (both significantly outperforming {\sc
  Cap3}).  Using the mouse fragment set used in Hazelhurst {\it et
  al.}  \cite{Hazelhurst08a} {\sc Peace} again shows slight
improvements in sensitivity and Type 1 error rates, with an 18\% speedup
for {\sc Peace} (see Supplementary Materials,
Section~\href{real_results}, for more details).

>>>>

\begin{abstract}
  We present {\sc Peace}, a standalone tool for the high-throughput
  {\it ab initio} clustering of transcript fragment sequences produced
  by Next Generation or Sanger Sequencing technologies.  Installed and
  run through an easily downloaded GUI, {\sc Peace} can process large
  data sets and group the fragments by gene association, achieving a
  greater sensitivity (without sacrificing runtime) than the competing
  clustering tool {\sc WCD}.  Through the GUI the user can collect
  cluster statistics and examine specific clusters for more
  comprehensive study.  {\sc Peace}, used as a ``pre-assembly'' tool,
  enables any assembly tool to work with individual small clusters of
  closely related sequences, resulting in a signficant: increases in
  sensitivity, robustness to sequenceing error, and reduction in
  overall runtime.
\end{abstract}

@@

Understanding an organism's transcriptome, the set of (spliced)
transcripts expressed by genes of the organism, is a vital step in
understanding the full functional and organizational role of the
genome in the life cycle of any eukaryote.  Studying the transcriptome
has led to gene discovery, provided information on splice variants,  
and helped shed light on the biological processes both controlling and
dictated by the genome \cite{Nagaraj07}.  But to access to those
transcripts, we must deal with the fragmented data produced by
both Next Generation and traditional Sanger sequencing technology.

Until recently, access to a transcriptome sequence was primarily
through the use of Expressed Sequence Tags (ESTs), single-pass cDNA
sequences derived from transcribed mRNAs and sequenced by Sanger
Sequencing technology.  More recently, Next Generation Sequencing
technology (NGS) is rapidly replacing Sanger Sequencing, allowing for
much more comprehensive coverage of the transcriptome.  For example,
ESTs in GenBank dbEST are increasingly the product of NGS technologies
such as 454 pyrosequencing, which enables the sequencing of novel and
rare transcripts at a considerably higher rate of coverage than did
Sanger Sequencing \cite{Cheung2006,Emrich2007}.  From a computational
perspective, this is a mixed blessing: while NGS provides immense
quantities of new information, it also provides immensely larger data
sets, requiring fast computational techniques if we are to process
them.

Given a set of transcript fragments sampled from across the genome, a
necessary first step in the analysis is that of clustering: separating
the fragments according to the transcript from which they were
derived.  Frequently performed implicitly by assembly tools,
clustering the data as a ``pre-assembly'' step has a number of
advantages.  Most significant among these: performing this step will
allow the application of the assembly tool to individual clusters --
saving significant amounts of time due to the reduced input size
\cite{Hazelhurst08a}.  But clustering is a computationally challenging
problem.  Even with the smaller number of ESTs produced using Sanger
sequencing, the runtime and memory requirements to cluster based on
pair-wise sequence alignments make such an approach infeasible.  The
much larger data-set size produced by NGS 
technologies exacerbates this problem.

Here we present {\sc Peace}, a tool for the {\it ab initio} clustering
of transcript fragments by gene, applicable to both NGS and
traditional Sanger Sequencing technologies.  Available through the
\href{http://www.peace-tools.org}{www.peace-tools.org} website, the
{\sc Peace} GUI allows the user to both easily install (locally or remotely)
and run the computational tool, as well as enabling transparent
parallel processing and providing various tools for result analysis.
Using a {\it minimum spanning tree} (MST) based clustering strategy
\cite{Jain99,Wan08} and the $d^2$ sequence distance function
\cite{Hide94}, {\sc Peace} generates the MSTs using a tailored version
of Prim's algorithm \cite{Prim57}.  The result is an easy-to-use and
clustering tool that , without any sacrifice in runtime, is
considerably more sensitive and robust to sequencing errors than other
clusters tools in the literature \cite{Burke99,Slater00,
  Huang99,Parkinson02,Kalyanaraman03,Malde03,Ptitsyn05,Hazelhurst08a, Picardi09}.

@@

\section{{\sc Peace}: Installation and Use}

The {\sc Peace} GUI, available from the {\sc Peace} website
(\href{http://www.peace-tools.org}{www.peace-tools.org}), handles the
details of both installation and job management and provides tools for
the initial analysis of the results.  By downloading and executing the
{\tt peace.jar} GUI Java program on a Linux, Windows or OS-X machine
running Java 1.6+, the user is presented with the {\sc Peace}
management tool.  Through this the user can install the computational
tool and perform a clustering of a data file in the FASTA format, view an initial analysis
of the clusters and produce files containing subsets of the clusters
for assembly by a tool such as {\sc Cap3} \cite{Huang99}.  A typical
(first) use of {\sc Peace} would generally require the the following
steps.

\noindent {\bf Tool Installation:} Installing the GUI requires only
that the Java file be downloaded from the website.  To install the
{\sc Peace} tool onto a local or remote machine, the user will select
from within the GUI the appropriate menu tab (Figure~1(a)), which
then starts the install wizard that will prompt for the appropriate
information.  Figure~1(b) illustrates the request for server
information; the user has chosen to install the {\sc Peace}
computational tool on a remote machine and is providing the necessary
connection information.  Server information is persistent between GUI
sessions, giving the user access to {\sc Peace} on the target machine
as needed.

\noindent {\bf Job Processing:} After importing the target sequence file into
the GUI, the user can start a new job and follow the wizard menus.
Figure~1(c) illustrates the process of specifying the number of
processors available (if running on a machine supporting the OpenMPI
protocol -- which will be determined during job installation).  Once
executed, the GUI will manage the job thread, alert the user when the
job is completed (or when the user next runs the GUI after completion)
and copy the final results back to the local machine if necessary.

\noindent {\bf Result Analysis:} Once the resulting clusters have been
computed, the user has several options
for analysis:
\begin{itemize}
\item {\bf Export:} The user can export the contents of one or more
  clusters into a FASTA format file, obtaining a subset of the original target file
  containing the sequences corresponding to the selected clusters
  ready for processing by an assembly tool (e.g. {\sc Cap3} \cite{Huang99}).
\item {\bf View Clustering}: The user may view a list of clusters,
  expanding selected clusters to a list of all individual sequences (illustrated
  in Figure~1(d)).
\item {\bf Classified Summary Graph:} The user may view a distribution
  of cluster sizes.  Further, the user may set up a {\it classifier},
  associating certain patterns with specific colors.  These patterns
  will be matched against the fragment header information from the
  original FASTA file, allowing the overlay of a colored cluster size
  distributions.  For example, if the sequence names contain unique
  string patterns denoting different cDNA libraries, the classifier
  can help the user to determine and visualize the differential expression profiles
  of different libraries for a given cluster.  The method of setting up a these classifiers, and the
  resulting histogram, is illustrated in Figure~1(e).
\end{itemize}
Extensive documentation for the tool has been posted on the {\sc
  Peace} website, as well as links to several tutorial videos demonstrating
{\sc Peace} use and capabilities.

@@

We have tested {\sc Peace} on both simulated and real data,
comparing results against those produced by the {\sc wcd}
clustering tool \cite{Hazelhurst08a} and the {\sc Cap3} assembly
tool \cite{Huang99} (which calculates a cluster as an implicit, intermediate
step).  For our simulation tests we used the {\bf ESTSim} tool
\cite{Hazelhurst03} to generate simulated transcript of varying length
fragments under different models
of error (Supplementary Materials, Section~C.1),
generating the fragments from the list of 100 zebra fish genes used in
the {\sc wcd} testing \cite{Hazelhurst08a}.  Tool parameters were
taken to match, as closely as possible, those used in the {\sc wcd}
study (see supplementary materials).  Our primary methods of quality
assessment were {\it sensitivity} (the fraction of fragment pairs from the
same gene that were correctly clustered together) and {\it Type 1
  error} (the fraction of genes that were divided between clusters)
\cite{Wang04,Hazelhurst08a}; other measurements are also discussed
in the supplementary materials.  In Figure~2(left) we see a
significant improvement of {\sc Peace} (blue) over {\sc wcd}
(green) and {\sc Cap3} (black) in its sensitivity to sequencing
errors, while in Figure~2(right) we see a comparable
improvement in Type 1 error.  In Figure~3 we look at
the number of singleton clusters (fragments not joined to any cluster),
which will not occur in our simulated sets; we again see significant
improvement in {\sc Peace}.  The sequential runtime of
{\sc Peace} is slightly improved over that of {\sc wcd} (see
Supplementary Materials, Figure~S4).

\begin{figure*}[b]
\centerline{
\includegraphics[trim=0cm 3cm 0cm 1cm, clip, scale=0.9]{screen.d/screen.pdf} 
}
\label{screen}
\caption{Screenshots of the {\sc Peace} GUI during execution, including (a)
GUI Welcome and server installation menu; (b) setup wizard for
installing the computational tool on a remote server; (c) execution
wizard for starting a selected job to be executed in parallel mode;
(d) basic cluster output; and (e) histogram view of cluster results
and classifier editor for setting up differential expression profiles.}
\end{figure*}

@@

In applying the tools to real data, we started with a set of
approximately $190,000$ clean Sanger ESTs derived from the {\it
  Chlamydomonas reinhardtii} genome \cite{Liang2008}.  To compute
sensitivity we used the {\it gmap} tool \cite{Wu05} to map the
fragment set to the genome, taking this as our reference clustering.
We see slight improvements in {\sc Peace} over {\sc wcd} in both
sensitivity and Type 1 error (both significantly outperforming {\sc
  Cap3}).  Using the mouse dataset used in Hazelhurst {\it et
  al.}  \cite{Hazelhurst08a} {\sc Peace} again shows slight
improvements in sensitivity and Type 1 error rates, with an 18\% speedup
for {\sc Peace} (see Supplementary Materials,
Section~\href{real_results}, for more details).

\section{Conclusions}

Here we have presented {\sc Peace}, a stand-alone tool for the
high-throughput clustering of transcript fragments, managed through a
user friendly GUI that enables both local and remote installation and
execution in sequential or parallel mode.  Based on a novel algorithm
for the clustering of the fragments by gene association, {\sc Peace}
shows significant improvement in sensitivity, without sacrificing runtime,
over the competing {\sc WCD} tool \cite{Hazelhurst08a}, and shows an
order of magnitude in improvement over the clustering preformed in the
course of assembly by the {\sc Cap3} tool \cite{Huang99}.

As a clustering tool based on sequence distance, {\sc Peace} faces
certain inherent limitations. For example, {\sc Peace} cannot handle
duplicate genes; like {\sc WCD}, it is unable to separate clusters
corresponding to genes with a greater than 88\% similarity.
Similarly, naturally biological effects such as the trans-splicing of
transcripts, or effects from poorly cleaned transcript data (e.g. the
failure to remove sequencing adapters), post-transcriptional
poly(A)/(T) tails, and low-complexity repeats can cause similar
effects in these clustering tools.  The problems can be handled
through the application of the assembler, and the ability to apply any
assembler to small cluster (as opposed to the data set as a while)
results in a significant reduction in overall assembly time.
